---
title: "F1 SQL analysis in R Studio"
author: "Maciej G"
output: html_document
---

The analysis is based on the Formula 1 World Championship (1950-2023) dataset, sourced from Kaggle. This relational dataset contains comprehensive records of races, drivers, constructors, qualifying results, and lap times, providing a rich foundation for complex SQL querying and performance analytics.

```{r}
library(DBI)
library(RSQLite)
library(readr)
```

I established a local ETL (Extract, Transform, Load) pipeline by importing multiple CSV datasets (drivers, results, races, etc.) into a temporary SQLite database in RAM. To ensure data integrity, I performed data type conversions, specifically formatting date and time strings to standard formats for accurate querying.

```{r}
con <- dbConnect(RSQLite::SQLite(), ":memory:")

drivers <- read_csv("Datasets/drivers.csv")
races <- read_csv("Datasets/races.csv")
results <- read_csv("Datasets/results.csv")
lap_times <- read_csv("Datasets/lap_times.csv")
circuits <- read_csv("Datasets/circuits.csv")
constructors <- read_csv("Datasets/constructors.csv")
constructor_results <- read_csv("Datasets/constructor_results.csv")

# Changing data types
drivers$dob <- as.character(drivers$dob)
lap_times$time <- as.character(lap_times$time)
races$date <- as.character(races$date)

# Coping data frames
dbWriteTable(con, "drivers", drivers, overwrite = TRUE)
dbWriteTable(con, "races", races, overwrite = TRUE)
dbWriteTable(con, "results", results)
dbWriteTable(con, "lap_times", lap_times, overwrite = TRUE)
dbWriteTable(con, "circuits", circuits)
dbWriteTable(con, "constructors", constructors)
dbWriteTable(con, "constructor_results", constructor_results)
```
**F1 dataset analysis**

Polish drivers sql to r table
```{sql, connection=con, output.var = "polish_drivers"}
SELECT 
    driverRef, 
    nationality 
FROM drivers 
WHERE nationality = 'Polish'
LIMIT 5
```

Polish F1 drivers
```{r}
print(polish_drivers)
```

TOP 5 drivers with most wins in MONACO GP
```{sql, connection=con}
SELECT d.forename, d.surname, COUNT(*) AS total_wins
FROM drivers d JOIN results res ON d.driverId = res.driverId
JOIN races r ON res.raceId = r.raceId
WHERE res.position = '1' 
  AND r.name = 'Monaco Grand Prix'
GROUP BY d.forename, d.surname
ORDER BY total_wins DESC
LIMIT 5
```

Fastest lap in F1 history
```{sql, connection = con}
SELECT d.forename, d. surname, l.time, l.milliseconds, r.name, c.name, r.year
FROM lap_times l 
  JOIN drivers d ON l.driverId = d.driverId
  JOIN races r ON l.raceId = r.raceId
  JOIN circuits c ON r.circuitId = c.circuitId 
ORDER BY l.time, l.milliseconds
LIMIT 1
```
Winners from grid position >15
```{sql, connection = con}
SELECT d.forename, d.surname, r.grid AS starting_position, rec.name AS GP_name, rec.year
FROM drivers d
  JOIN results r ON d.driverId = r.driverId
  JOIN races rec ON r.raceId = rec.raceId
WHERE r.positionOrder = 1 AND grid > 15
ORDER BY starting_position DESC
```
Which era driver raced in?
```{sql, connection = con}
SELECT forename, surname, dob,
  CASE
    WHEN dob < '1950-01-01' THEN 'Pioneer'
    WHEN dob BETWEEN '1950-01-01' AND '1980-01-01' THEN 'Classic'
    ELSE 'Modern'
  END AS Era
FROM drivers
ORDER BY dob DESC
LIMIT 5
```
Number of drivers form each era
```{sql, connection = con}
SELECT
  CASE
    WHEN dob < '1950-01-01' THEN 'Pioneer'
    WHEN dob BETWEEN '1950-01-01' AND '1980-01-01' THEN 'Classic'
    ELSE 'Modern'
  END AS era, 
COUNT(*) AS total
FROM drivers
GROUP BY era
ORDER BY total DESC
```
Number of drivers from different countries (>50)
```{sql, connection = con}
SELECT nationality, COUNT (*) AS total
FROM drivers
GROUP BY nationality
HAVING total > 50
ORDER BY total DESC
```
Points by team after 2010, only 1000+
```{sql, connection = con}
SELECT c.name, SUM(cr.points) AS total_points
FROM constructors c 
  JOIN constructor_results cr ON c.constructorId = cr.constructorId
  JOIN races r ON cr.raceId = r.raceId
WHERE r.year > 2010
GROUP BY c.name
HAVING SUM(cr.points) > 1000
ORDER BY 2 DESC
```
CTE, Window Functions

Each GP winner in year 2023
```{sql, connection = con}
WITH position_table AS (
  SELECT d.surname, r.name, res.points, r.year,
    RANK() OVER (PARTITION BY r.raceId ORDER BY res.points DESC) as position
  FROM results res
  JOIN drivers d ON res.driverId = d.driverId
  JOIN races r ON res.raceId = r.raceId
)
SELECT surname AS driver, name AS gp_name, points, year, position
FROM position_table 
WHERE position = 1 AND year = 2023
```
Verstappen 2023 GP points + difference to previous one
```{sql, connection = con}
WITH ver_points_2023 AS (
  SELECT 
    r.name, 
    r.date,
    res.points as current_points,
    LAG(res.points) OVER (ORDER BY r.date) as previous_points
  FROM results res
  JOIN drivers d ON res.driverId = d.driverId
  JOIN races r ON res.raceId = r.raceId
  WHERE d.surname = 'Verstappen' AND r.year = 2023
  ORDER BY r.date
)
SELECT *, (current_points - previous_points) AS points_diff
FROM ver_points_2023
```
H2H Ferrari drivers, position in 2023 GPs
```{sql, connection = con}
SELECT 
  r.name AS GrandPrix,
  d1.surname AS Driver_A,
  r1.positionOrder AS Pos_A,
  d2.surname AS Driver_B,
  r2.positionOrder AS Pos_B,
  (r2.positionOrder - r1.positionOrder) AS position_gap
FROM results r1
JOIN results r2 ON r1.raceId = r2.raceId AND r1.constructorId = r2.constructorId
JOIN drivers d1 ON r1.driverId = d1.driverId
JOIN drivers d2 ON r2.driverId = d2.driverId
JOIN races r ON r1.raceId = r.raceId
WHERE r1.constructorId = 6
  AND r.year = 2023
  AND d1.surname > d2.surname
```
Number of wins by driver nationality
```{sql, connection = con}
SELECT d.nationality, COUNT(CASE WHEN r.position = 1 THEN 1 END) AS wins
FROM drivers d 
  LEFT JOIN results r ON d.driverId = r.driverId
GROUP BY nationality
ORDER BY 2 DESC
```

```{r}
dbDisconnect(con)
```




